Keep your Laziness in Check

### 2018-01-11 paper/main.tex
rewriting intro

\begin{itemize}
\item Laziness is a fundamental feature in Haskell \leo{WhyFPmatters?}
\item However, reasoning about programs with laziness (performance, behavior) is hard.
\item It is very easy to use a too-lazy or too-strict implementation that yields different results.
\item FRP quote here?
\end{itemize}

Consider the following implementation of the \lk{sum} function:
\leo{I created an environment for this so that we can tweak them all as needed}
\begin{inlinecode}
sum = foldl (+) 0
\end{inlinecode}
%
\begin{itemize}
\item As many readers/Haskellers have probably encountered, such an implementation
   can be inefficient due to space leaks because of laziness.
\item When summing over a large list, instead of calculating the running sum as we traverse
   the list, we build an increasingly larger thunk that will be forced in the end.
\item An easy solution is to use a stricter fold, like \lk{foldl'}, which forces the
   fold accumulator in weak head normal form.
\item However, even that is not enough for all cases.
\item Consider a single-traversal computation of an average where we keep both a running sum and a running length in an accumulator.
\end{itemize}
%
\begin{inlinecode}
avg = uncurry (/) . foldl (\(s,n) x -> (s + x, n + 1)) (0,0)
\end{inlinecode}
%
\begin{itemize}
\item In this case, both \lk{foldl} and \lk{foldl'}, construct a thunk, since forcing a pair into weak head normal form does not force either component.
\item We need an implementation that is strict in those components as well.
\item Which leads to the natural question, how can we test the laziness behavior of functions?
\item Even worse, how can we even {\em specify} the laziness behavior?
\end{itemize}

\leo{We probably need a paragraph that explains what we want to do here, not directly a jump into the contributions}

Our technical contributions are as follows:
\begin{itemize}
\item We introduce a novel approach to specifying the laziness behavior of functions that is
{\em precise} \leo{define}. \leo{Also add drawbacks/partiality here? If not here, where in the intro?}
\item We propose a random testing approach to test such specifications.
\item We implement StrictCheck. \leo{In order for these last two points to be different, we need
something about the cool/awesome hacks in the implementation to point out here}
\item As a side contribution? We fix the function generaton of QuickCheck to produce lazy/partial functions instead
of fully strict ones all the time.
\end{itemize}
%
Section~\ref{sec:related} discusses related work. We conclude and draw
directions for future work in Section~\ref{sec:concl}.

>>>>

\hzh{Rewrite counter: 1}
\leo{I want to keep a counter on how many times we'll rewrite this intro :)}

John Hughes has argued adequately that lazy evaluation brings tremendous power
to modularization of functional programming in his seminal work \textit{Why
Functional Programming Matters} ~\cite{WhyFPMatters}. Haskell is one of the most
popular programming languages that gives its programmers the power of lazy
evaluation.
%
Unfortunately, reasoning about the performance of lazy programs in Haskell is
difficult. Programmers often accidentally make a program too lazy or too strict,
and as a result, their programs behave differently than expected.

\leo{Didn't we have an FRP quote by someone that Laziness was the third source of
bugs or something? Work that in?}
\hzh{Ryan Trinkle (author of Reflex) said laziness was the 3rd most difficult thing to
get right, but let's check with him on whether we can quote him on that.}

For example, consider the following simple Haskell implementation of the \lk{sum} function:
\leo{I created an environment for this so that we can tweak them all as needed}
\begin{inlinecode}
sum = foldl (+) 0
\end{inlinecode}
%
Although this seemingly innocent implementation of sum will produce correct
results on small lists of numbers, it will quickly run out of memory on larger
lists. The culprit here is the laziness behavior of \lk{foldl}---it does not
force the accumulated result during the fold. Instead, calling this \lk{sum} on
a list produces a thunk that is linear in the size of of the input list, and
this thunk will not be evaluated until the result of sum is used in another
computation.
%
We can fix this space leak for \lk{sum} using a strict fold, \lk{foldl'}, which
eagerly forces the accumulator, and avoids creating a space-leaking thunk.

However, using \lk{foldl'} is not sufficient to avoid space leaks in
general. Consider a single-traversal computation of an average where we keep
both a running sum and a running counter in an accumulator.
%
\begin{inlinecode}
avg = uncurry (/) . foldl (\(s,n) x -> (s + x, n + 1)) (0,0)
\end{inlinecode}
%
In this case, both \lk{foldl} and \lk{foldl'} construct thunks linear in the
size of the input list, since forcing a lazy pair only evaluates it to its weak
head normal form, and it does not evaluate either component of the pair.

How, then, can we test and catch these laziness bugs before they
manifest themselves in production? Is it possible to precisely specify
laziness behavior just as we would for functional correctness?
%
In this paper, we demonstrate techniques for observing laziness in Haskell, and
propose a system of creating specifications of lazy Haskell programs. We
implement these ideas as a Haskell library.

Our technical contributions are as follows:
\begin{itemize}
\item We introduce a novel approach to specifying the laziness behavior of functions that is
{\em precise}. Such specifications are expressive enough to capture exactly how
much evaluation occurs within a data structure \leo{Also add
drawbacks/partiality here? If not here, where in the intro?} \hzh{I'm not sure
if we should bring in partiality (in the sense of divergence/nontermination) yet
since that seems like a rather technical detail of the spec}
\item We propose a framework for testing against such specifications.
\item We improve upon the random function generators from QuickCheck to generate functions with arbitrary laziness behaviors.
\item We demonstrate non-trivial generic programming techniques to apply StrictCheck to almost all data types.
\end{itemize}
%
Section~\ref{sec:related} discusses related work. We conclude and draw
directions for future work in Section~\ref{sec:concl}.
